Learning how to act: making good decisions with machine learning

### 2017-09-02 thesis/offcuts_chapter6.tex
adds intro to causal-bandit chapter
adds abstract, improves contributions section

\chapter*{Todo}
\begin{enumerate}
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique? It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition. 
\end{enumerate}

@@

\section{careful prior}

\subsubsection{Example}

\eqn{
U &\sim N(0 , 1)\\
Z &\sim N(2U , 0.09)\\
X &\sim N(0.5Z , 1)\\
Y &\sim N(3U + -Z + 0.5X , 0.25)
}

\begin{figure}[h]
\includegraphics[scale=.3]{prior_no_prior}
\caption{no prior, posterior on $w_{yx}$ centered around $w_{yx}$}
\end{figure}

\begin{figure}[h]
\includegraphics[scale=.3]{prior_bad_prior}
\caption{Prior around $w_{yz}$, $\beta \sim N(w_{yz},\sigma = 0.5)$. The posterior on $w_{yx}$ is biased away from its true value}
\end{figure}

@@

\section{intro}
Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches. They do not contain novel technical results but present the material in a unified way. Chapter \ref{chap:causal-bandit} is a novel contribution. It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.

>>>>

\section*{Questions for Tor}
- How do I fix the page numbering in my thesis?
- How does one derive the regret decompostion
- Any suggestions on reviewers for my thesis (or people who should not be reviewers)
- Who else do I need to cite in my bandits chapter?

\chapter*{Todo}
\begin{enumerate}
\item Check what I actually have to do to submit
\item Add something about Granger causality - does this belong in chapter 2 or chapter 3?
\item Sell my PhD as two key contribution. 1) A technical one, the other educational and unifying (across viewpoints)
\item Add a chapter in 'PhD prose' making some statement about the intersection of prediction and causality. When is a problem causal, what bounds does predictive accuracy give on causal models? To what degree is predictive accuracy a metric on which causal models can be judged? 
\item Benchmark estimation algorithms
\item clarify causal interpretation of Dirichlet models
\item Empirical evaluation of bandit algorithms (updated to include latest suggestions). Maybe a proper software bench marking solution?
\item If comparing bandit performance, it would be interesting to include some that violate the stochastic bandit assumption. For example, if the rewards change slowly over time or jump suddenly a limited number of times, or with some small probability. 
\item Does defining causality in term of the truncated product formula ensure that a causal Bayesian network is unique? It would seem no, one can ad links (although this is hardly useful) - seems like there is no reverse condition. 
\item Statistisian/Econometricians and computer science seem to use different definitions of optimality (bandits vs efficient estimators). Clarify this - would make a good blog post I think. 
\end{enumerate}

@@

@@

\section{intro}
\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

We study the problem of using causal models to improve the rate at which good interventions can be learned online in a stochastic environment. 
Our formalism combines multi-arm bandits and causal inference to model a novel type of bandit feedback that is not exploited by existing approaches.
We propose a new algorithm that exploits the causal feedback and prove a bound on its simple regret that is strictly better (in all quantities) 
than algorithms that do not use the additional causal information.

Chapters \ref{chap:causality} and \ref{chap:interventionalist_view} predominantly review the existing literature with the goal of highlighting connections between the different viewpoints and approaches. They do not contain novel technical results but present the material in a unified way. Chapter \ref{chap:causal-bandit} is a novel contribution. It formally connect causal graphical models and multi-armed bandit problems, demonstrates this leads to a form of structure that had not previously been analysed within the bandit literature and describes an algorithm that can leverage this structure to make better decisions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

A secondary goal is to summarise the key strands of causal inference research from economics and social sciences for the machine learning community and highlight connections and productive areas for future research.

I describe the key existing frameworks for

Demonstrate connections between existing lines of work within the bandit and casusal inference communities.

To summarise the key strands of causal inference research from economics and social sciences for the machine learning community.
\section{Random}

### 2017-09-02 thesis/thesis_map.tex
adds abstract, improves contributions section

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib} %numbers
\usepackage{wrapfig}
\usepackage{centernot}

@@

\maketitle

@@

@@

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Image classification
\item Forecasting the weather
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Customer churn modelling
\item Demand prediction for inventory control
\item Playing Go 
\end{itemize}

@@

\section{An overview of this thesis and its contributions}

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields these techniques all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

@@

The goal of this thesis is to draw together the key approaches to solving causal problems from both the interventionalist and observational viewpoints. In chapter \ref{chap:causality} I describe key frameworks for causal inference from observational data, discuss how the relate to one-anther and review how they can be leveraged to answer causal queries. Chapter \ref{chap:interventionalist_view} covers the interventionalist viewpoint, including traditional randomised experiments and the closely related multi-armed bandit problems. It describes the key problems and results within the bandit literature and highlights some interesting connections between existing strands of work on bandits and causal inference. In chapter \ref{chap:causal-bandit}, I develop a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. The causal models allow us to represent our knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

@@

@@

\subsection{Defining causal effects} 
So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data. This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set. From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.

@@

\section{Randomised experiments}
Randomised controlled trials are often presented as the gold standard for determining causal effects. What is it about randomisation that makes it so important when it comes to causality? The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}. If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process. Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.

@@

\subsection{Stochastic bandits: Approaches and results}

@@

\subsection{Pure-exploration problems}
Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}. In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$. The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.

@@

\subsection{Adversarial Bandits}

@@

\subsubsection{The causal structure of contextual bandits}

@@

\subsection{Learning from logged bandit data}
Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}. In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$. The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.

@@

\subsection{Adding structure to actions}

@@

\section{Causal bandits with post action feedback}

@@

\subsection{Experiments}
We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions. Thomson sampling and UCB are optimised to minimise cumulative regret. We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean. The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.

@@

>>>>

\documentclass[11pt,a4paper,twoside]{report}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib} %numbers
\usepackage{wrapfig}
\usepackage{centernot}

@@

\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

        \Huge
        {Learning how to act: making good decisions with machine learning}

        \vspace{2cm}
        \Large
        {Finnian Lattimore}

        \vfill
        \Large
        A thesis submitted for the degree of Doctor of Philosophy

         The Australian National University

        \vspace{1cm}

        September 2017

    \end{center}
\end{titlepage}

@@

\clearpage
\vspace*{\fill}
\begin{center}
\begin{minipage}{.9\textwidth}
Except where acknowledged in the customary manner, the material 
presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
university.

\vspace{40mm}  % vertical space

\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
\hspace{80mm} Finnian Lattimore\par
\hspace{80mm} September, 2017
\end{minipage}
\end{center}
\vfill % equivalent to \vspace{\fill}
\clearpage

\begin{abstract}

This thesis is about machine learning and statistical approaches to decision making. How can we learn from data to anticipate the consequence of, and optimally select, interventions or actions? Such problems, from deciding which medication to prescribe to patients, who should be released on bail or how much to charge for insurance, to designing policies to reduce unemployment or improve school attendance are ubiqutes and have far reaching impacts on our lives. There are two fundamental approaches to learning how to act: reinforcment learning, in which an agent directly intervenes in a system and learns from the outcome, and observational causal inference, whereby we seek to infer the outcome of an intervention from observing the system.

The goal of this thesis to to connect and unify these key approaches. I introduce causal bandit problems: a framework which combines causal graphical models, which were developed for observational causal inference, with multi-armed bandit problems, which are a subset of reinforcment learning problems that are simple enough to admit formal analysis. I show that knoweledge of the causal structure allows us to transfer information learned about the outcome of one action to predict the outcome of an alternate action, yielding a novel form of structure between bandit arms that cannot be expoited by existing algorithms. I propose an algorithm for causal bandit problems and prove bounds on the simple regret demonstrating it is close to mini-max optimal and strictly better than algorithms that do not use the additional causal information.

\end{abstract}

%\chapter*{Acknowledgements}

@@

\begin{itemize}
\item Speech recognition (for systems like Siri or Google Assistant)
\item Image classification
\item Forecasting the weather
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Customer churn modelling
\item Demand prediction for inventory control
\item Playing Go 
\end{itemize}

@@

\section{Observe or intervene: two very different approaches to causal problems}

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

@@

\section{This thesis and its contributions}

I view causal problems as one of the greatest arising challenges for machine learning. They incorporate a large set of problems, of huge practical significance, that require us to go beyond pattern recognition, but are well short of general artificial intelligence. For the major advances in representation and pattern recognition developed within machine learning to be effectively applied in many areas of medicine, economics, social sciences and industry we need to understand how to leverage our improved approaches to prediction to tackle causal problems.

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework which unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. My other contributions are primarily connective rather than technical. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits} I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation} I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of of causal effect estimation from observational data. Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.

In chapter \ref{chap:causal-bandit} I introduce causal bandit problems: a framework which unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback} I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret} I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general} I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion} I discuss extensions and potential future work.

@@

\todo{is there a mechanical method to go from one framework to another}

\todo{does it matter if variables are discrete or continuous}

@@

\subsection{Defining causal effects} 
\label{sec:defining-causal-effects}
So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data. This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set. From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.

@@

\section{Randomised experiments}
\label{sec:randomized_experiment}
Randomised controlled trials are often presented as the gold standard for determining causal effects. What is it about randomisation that makes it so important when it comes to causality? The graphical model for a randomised controlled experiment is shown in figure \ref{fig:random_experiment_network}. If we assume perfect compliance (everyone takes the treatment that we select for them) then we have a perfect model for the treatment assignment process. Since treatment is assigned randomly, there can be no other variables that influence it and thus no confounding variables that effect both treatment and outcome.

@@

\subsection{Stochastic bandits: Approaches and results}
\label{sec:stochastic-bandit}

@@

\subsection{Pure-exploration problems}
\label{sec:pure-exploitation-problems}
Another problem that has attracted a lot of recent attention \citep{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the stochastic multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}. In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$. The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm.

@@

\subsection{Adversarial Bandits}
\label{sec:adversarial-bandits}

@@

\subsubsection{The causal structure of contextual bandits}
\label{sec:causal-structure-contextual-bandit}

@@

\subsection{Learning from logged bandit data}
\label{sec:off-policy-evaluation}
Another topic of interest within the bandit community, which is deeply connected to causal effect estimation from observational data, is learning from logged bandit feedback data or off-policy evaluation\citep{Langford2008a,strehl2010learning,Li2010,Dudik2011,Bottou2013,Swaminathan2015}. In this setting, the learner has a data set $S = \set{(\vec{X}_t,A_t,\cfb{Y}{A_t})}_{t=1}^T$, which is assumed to have been generated by a stochastic contextual bandit environment interacting with some unknown, potentially stochastic, policy $\pi(\vec{x}_t,h_t)$, where $h_t$ is the sequence of observed data up to time $t$. The goal of the learner is to evaluate the value of an alternate policy, $\pi'$, for selecting actions, often with the underlying motivation of identifying an optimal policy within some space of policies $\Pi$.

@@

\subsection{Adding structure to actions}
\label{sec:adding-structure-bandit-actions}

@@

\section{Causal bandits with post action feedback}
\label{sec:causal-bandit-post-action-feedback}

@@

\subsection{Experiments}
\label{sec:causal-bandit-experiments}
We compare Algorithms \ref{alg:simple} and \ref{alg:general} with the Successive Reject algorithm of \citet{Audibert2010}, Thompson Sampling and UCB under a variety of conditions. Thomson sampling and UCB are optimised to minimise cumulative regret. We apply them in the fixed horizon, best arm identification setting by running them up to horizon $T$ and then selecting the arm with the highest empirical mean. The importance weighted estimator used by Algorithm \ref{alg:general} is not truncated, which is justified in this setting by Remark \ref{rem:truncate}.

@@

Causal questions are so important

Summarise the amazing stuff I have done

Many interesting future directions for research
